---
title: "Getting started with logrittr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with logrittr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## Motivation

In SAS, every DATA step prints a log that tells you exactly what happened:

```
NOTE: There were 120000 observations read from WORK.SALES.
NOTE: 7153 observations were deleted.
NOTE: The data set WORK.SALES has 112847 observations and 11 variables.
```

R's `dplyr` pipelines are silent by default. `logrittr` fills that gap with
`%>=%`, a pipe operator that logs at each step:

- row count before/after (with signed delta)
- column count before/after (with signed delta)
- column names added or dropped
- elapsed time

No function masking, no dependencies beyond `cli` and `stringr` for coloring and formatting in console.

> **Tip**: Fira Code users: with ligatures enabled, `%>=%` renders as a single
> wide arrow, close from a regular pipe.

## Installation

```{r install}
install.packages("logrittr", repos = "https://guillaumepressiat.r-universe.dev")

# alternatively
# remotes::install_github("GuillaumePressiat/logrittr")
```

## Basic usage

```{r basic}
library(logrittr)
library(dplyr)

iris %>=%
  as_tibble() %>=%
  filter(Sepal.Length < 5) %>=%
  mutate(rn = row_number()) %>=%
  group_by(Species) %>=%
  summarise(n = n_distinct(rn))
```

```
i as_tibble()                               rows:       150 +0        cols:    5 +0    [   0.0 ms]
i filter(Sepal.Length < 5)                  rows:        22 -128      cols:    5 +0    [   1.0 ms]
i mutate(rn = row_number())                 rows:        22 +0        cols:    6 +1    [   0.0 ms]
  added: rn
i group_by(Species)                         rows:        22 +0        cols:    6 +0    [   0.0 ms]
i summarise(n = n_distinct(rn))             rows:         1 -21       cols:    2 -4    [   1.0 ms]
  dropped: Sepal.Length, Sepal.Width, Petal.Length, Petal.Width, rn
  added: n
```

`%>=%` is fully composable with `|>` and `%>%`. Use it only where you want
visibility, and fall back to the native pipe for the rest.

## Nested pipelines

When `%>=%` appears inside an argument (e.g. inside `semi_join()`), the nested
steps are automatically indented with a `>` prefix so they are visually
distinct from the main pipeline:

```{r nested}
iris %>=%
  as_tibble() %>=%
  filter(Sepal.Length < 5) %>=%
  mutate(rn = row_number()) %>=%
  semi_join(
    iris %>% as_tibble() %>=%
      filter(Species == "setosa"),
    by = "Species"
  ) %>=%
  group_by(Species) %>=%
  summarise(n = n_distinct(rn))
```

```
i as_tibble()                                    rows:       150 +0        cols:    5 +0    [   0.0 ms]
i filter(Sepal.Length < 5)                       rows:        22 -128      cols:    5 +0    [   1.0 ms]
i mutate(rn = row_number())                      rows:        22 +0        cols:    6 +1    [   0.0 ms]
  added: rn
i   > filter(Species == "setosa")                rows:        50 -100      cols:    5 +0    [   0.0 ms]
i semi_join(iris %>% as_tibble() %>=%            rows:        20 -2        cols:    6 +0    [   5.0 ms]
  filter(Species == "setosa"), by = "Species")
i group_by(Species)                              rows:        20 +0        cols:    6 +0    [   0.0 ms]
i summarise(n = n_distinct(rn))                  rows:         1 -19       cols:    2 -4    [   1.0 ms]
  dropped: Sepal.Length, Sepal.Width, Petal.Length, Petal.Width, rn
  added: n
```

## Options

All display options are controlled via `logrittr_options()`:

```{r options}
logrittr_options()
#> $wrap_width
#> [1] 52
#> $big_mark
#> [1] " "
#> $lang
#> [1] "en"
#> $max_cols
#> [1] 5
```

### Language

Switch to French with `lang = "fr"` (the metrics line uses `lignes`
instead of `rows`):

```{r lang}
logrittr_options(lang = "fr")

iris %>=%
  select(Species, Sepal.Length, Sepal.Width) %>=%
  filter(Sepal.Length > 5)
```

```
i select(Species, Sepal.Length, Sepal.Width)  lignes:       150 +0       cols:    3 -2    [   0.0 ms]
  dropped: Petal.Length, Petal.Width
i filter(Sepal.Length > 5)                    lignes:       118 -32      cols:    3 +0    [   0.0 ms]
```

### Thousands separator

```{r bigmark}
logrittr_options(lang = "en", big_mark = ",")

big <- data.frame(x = seq_len(1e6), y = rnorm(1e6))
big %>=% filter(x > 500000)
```

```
i filter(x > 500000)          rows:   500,000 -500,000  cols:    2 +0    [  45.0 ms]
```

### Column name truncation

When a `select` or `join` adds or drops many columns, only the first `max_cols`
names are shown to keep the log readable:

```{r maxcols}
logrittr_options(max_cols = 2, lang = "en")

iris %>=%
  as_tibble() %>=%
  select(Species, Sepal.Length)
```

```
i select(Species, Sepal.Length)             rows:       150 +0        cols:    2 -3    [   0.0 ms]
  dropped: Sepal.Width, Petal.Length, and 1 other
```

Use `max_cols = Inf` to always display all names:

```{r maxcols_inf}
logrittr_options(max_cols = Inf)
```

### Restoring defaults

`logrittr_options()` invisibly returns the previous values, which makes it easy
to restore the state after a temporary change:

```{r restore}
old <- logrittr_options(lang = "fr", big_mark = ",")
# ... work ...
do.call(logrittr_options, old)  # restore previous state
```
